Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila

### 2018-09-28 phil_paper/jmp_wjv_v2.tex
try splitting best-response

%  Shared connections record high levels of monthly consumption, which can expose them to high prices when marginal prices increase with usage . 
This paper analyzes the welfare impacts of pricing policies in the presence of informal water sharing.  Welfare impacts depend not only on heterogeneity in demand for water across the population, but also on costs associated with forming and maintaining water sharing relationships.  To measure these features, I merge administrative data covering 1.5 million water connections in Manila from 2010 to 2015 with detailed survey data on the number of households and people using each water connection.  With comprehensive geographic coverage of half of Manila and a variety of price variation, these data allow for a rich characterization of water demand.  Additional variation in water source choices identify the relative fixed and marginal costs associated with sharing water with a neighbor.  I find that despite relatively high marginal costs of sharing, shared connections allow the city to save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households at low marginal prices.  At the same time, informal sharing networks act as resale markets for water, undermining the extent to which cities can redistribute surplus through increasing marginal prices.

@@

Combining the model and estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,wsup,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensuring that the water provider exactly covers costs.  This policy reduces access by incentivizing households to substitute toward water vendors (which mainly include deepwells and water refilling stations).

@@

In focusing on water sharing behavior, this approach abstracts away from several features that may also affect optimal pricing policy.  First, I do not directly model environmental, health, or other externalities associated with piped water use.  From a policy perspective, water shortages may lead policymakers to promote conservation through higher marginal prices while positive health externalities may motivate further subsidizing piped water (\cite{timmins2002measuring}, \cite{galiani2005water}).\footnote{\cite{bennett2012does} shows that piped water can also lead to negative externalities through worse sanitation, possibly motivating higher prices for piped water.}  Second, I assume that government regulator is able to fully observe production costs for the water provider, eliminating scope for prices to correct for possible principal-agent frictions as characterized by \cite{laffont1993theory}.  Finally, the counterfactual exercises borrow the public-private partnership regulatory framework in Manila by assuming that revenue must fully cover production costs.  Additional data on the costs of raising public funds and risks of regulatory capture would allow for a more flexible analysis taking into account the possibility of public funding for water (\cite{laffont2005regulation}).

@@

%Monthly quantity choices across different households and at different prices reveal information about household demand.  Observing multiple quantities choices for the same household allows for a rich characterization of demand heterogeneity at the household-level.  
Evaluating different pricing policies requires measuring demand at the household level as well as costs associated with providing piped water.  For information about these attributes, I developed a partnership with one of the regulated, private providers in Manila who provided both billing records as well as information about the regulatory structure, pricing, and production costs.  Complete billing records for each water connection including the monthly meter reading, the total bill charged, and the precise location for each meter serve as the primary measure of quantity choices.  These data span 2010 to 2015, during which connections grew from 900,000 to 1.5 million water connections, generating nearly 100 million transaction records.  Connections are divided into residential (90\%), semi-business (4\%), commercial (5\%), and industrial (1\%).  This paper focuses on residential and semi-business connections.\footnote{Semi-business connections are largely composed of households that own small roadside stands.  These connections provide a useful source of price variation which is described in Appendix~\ref{appendix:ratechangeelasticity}.}  The monthly meter reading is used as the measure of consumption for each connection.  Meter readings are rounded to the nearest cubic meter, and I limit consideration to meter readings below 100 $\text{m}^{3}$ in order to exclude pipe leaks as well as meter reading errors (readings above 100 $\text{m}^{3}$ compose less than 0.7\% of the sample).

Since the object of interest for welfare is household-level demand, I merge the billing records to unique survey data recording information on households using each connection.  To evaluate the performance of the water provider, the regulator contracted with a local university to conduct the Public Assessment of Water Services (PAWS) in three waves from 2008 to 2012, surveying around 50,000 water connections.\footnote{Appendix~\ref{appendix:pawssampling} describes the sampling design and tests the extent to which it may be correlated with demographic characteristics.  This exercise finds relatively weak correlations in economic terms with key characteristics.}  I have collected unique identifiers for this survey, which allows responses to be linked to water connections.  The connection survey records the total number of households and people using each connection, providing the primary measure of sharing behavior.  The connection survey also includes basic demographics for the owner of the connection.  The demographics used in this paper are household size, dwelling type (whether apartment, duplex, or single house), number of employed household members, and whether the head of household is employed in a low-skill profession.
%consumption and water source choices at the household-level.  
%To examine both consumption decisions and water source choices at the household-level, I construct a dataset that merges (1) monthly water usage for around 50,000 connections, (2) demographics and water sharing measures for households using these connections, and (3) a cross-section of all households and their water source choices.  Put together, these data form a representative sample of households in half of Metro Manila.

A full analysis of water usage in Manila also requires describing households that choose to use water from local vendors.  For information on households that are using water from alternative sources, I turn to the 2010 Census of Population and Housing.  Households are defined as connected to the piped network if they report using a ``faucet community water system'' for their cooking, laundry, or bathing needs.\footnote{The census was designed largely for rural populations so it does not directly ask whether households are connected to a piped water network.}  Alternatively, I define households as using from a water vendor if they report any other category, which includes deep wells, peddlers, or other sources.  I then construct similar demographic measures to the connection survey using questions in the census and compare these measures across these data sources in Table~\ref{table:pawstocensus} of Appendix~\ref{appendix:pawstocensus}.  Households that own a connection in the connection survey are similar to households connected to the piped water network in the Census data.  To construct a profile of water source choices for the full population, I randomly sample from the pool of households using from vendors in 599 wards of the city so that the data form a representative sample when combined with households covered by the connection survey in each small area.
%\footnote{Regions are chosen to overlap with ward administrative/political boundaries while ensuring that I am able to obtain a large enough sample of vendor households for each area to perform the counterfactual analysis in Section~\ref{section:fixedcostestimation}.}

For additional information on water sharing relationships between households, I fielded a small, non-representative survey of 600 households via mobile phone app.  This survey asks about issues associated with using different water sources, basic demographics, as well as how the bill and connection fee are paid for sharing households.  Appendix~\ref{appendix:pollfish} includes additional details on this survey and finds roughly similar measures between the connection survey and the mobile survey, suggesting that the mobile survey may provide a reasonable descriptive picture of sharing in Manila.

The water provider also maintains call records of all service complaints, which can be linked to individual connections.  This data is used to measure instances where households face unanticipated shocks to their water source as part of the identification strategy to estimate the marginal costs of sharing water.  Finally, the water provider tracks data on the marginal cost of pumping an additional cubic meter of water as well as the costs associated with installing new connections.  These data provide inputs into the provider's budget constraint, which are necessary for computing counterfactual pricing regimes.  Supplemental survey data from a subset of Manila include income measures that are used to impute income for the remainder of the sample across a variety of demographic variables and their interactions.

@@

The welfare consequences of informal sharing networks depend on both the types of households that choose to share with neighbors as well as the costs of these networks.  To provide some descriptive evidence, Table~\ref{descriptives} merges the connection survey of water connections with billing records associated with each connection.  The columns group connections by the number of households they serve: one, two, or at least three.

The first rows provide demographic information from the connection survey for households who own connections.  Smaller and younger households are more likely to share water with neighbors.  Because household size is a strong proxy for water demand, this finding suggests that sharing a connection is negatively correlated with household water demand.  Households with lower imputed income more commonly share water connections.  Sharing households are over twice as likely to reside in duplexes possibly because transporting water between neighbors requires less effort for duplex residents.

@@

Table~\ref{descriptives} summarizes water use and sharing behavior for these connections, merging information from the connection survey with the billing records.  Larger numbers of households sharing a water connection are correlated with higher total water usage, but less water per person.\footnote{Total People Served by a connection includes both the owning household and any other households using a connection.}  This decline in usage per person may have many possible drivers: (1) sharing households face higher prices by locating at a higher point on the tariff schedule (Figure~\ref{figure:tariff}), (2) sharing households may face hassle costs in fetching water from neighboring taps, (3) these price effects may incentivize low-demand households to select into sharing relationships, consistent with the demographic patterns, and (4) low-demand households may have idiosyncratic preferences for sharing relationships.  On average, monthly water bills (inclusive of service fees) take up around \input{tables/incshr}\unskip\% of household income, which does not include other unobserved costs in accessing water.  These findings suggest that policies which affect sharing choices can have significant welfare consequences.

%Also, the mobile survey indicates that only around 30\% of sharing households would substitute to toward a neighbor's connection if they no longer had access to their primary water source.  For this reason, I treat connections reported to serve more than three households as only serving three households.  
Table~\ref{descriptives} also computes the total number of households using from each source.  I assume that households sharing with one or two other households are uniquely linked to one connection so that the population of households connected to the network can be computed by summing the number of households served by each connection.  This assumption is consistent with the presence of fixed costs in setting up a connection or a sharing relationship.  To the extent that households are drawing water from multiple connections, this assumption may lead to an overestimation of the population engaged in sharing relationships.  Under this assumption, I find that \input{tables/shr_individual}\unskip\% of connected households receive their water from an individual connection, \input{tables/shr_1hh}\unskip\% are involved in a sharing relationship with a single neighbor, and \input{tables/shr_2hh}\unskip\% are in a sharing relationship with 2 or more other households.  These proportions do not take into consideration households that are using from water vendors, which are discussed in more detail below.  Each water connection contributes over six years of billing records on average.

Since the connection survey only includes households connected to piped water, I use census data to learn about households using from water vendors.  \input{tables/vendor_total}\unskip\% of the population report using vendors, which is composed of \input{tables/vendor_well}\unskip\% of households using from a ``deep well'' and \input{tables/vendor_peddler}\unskip\% using from a ``peddler'' while the remaining fall into an ``other'' category.  Anecdotal evidence suggests that groundwater from wells often has very low quality.  Appendix Table~\ref{table:pawstocensus} compares means of demographic variables between connection owners from the connection survey and all households from the census.  Demographic variables are similar between connection owners in the connection survey and the census data.  Households using from vendors have lower imputed incomes, greater probability of low-skill employment, and slightly fewer members than own-use households.  Across most dimensions, households using from vendors more closely resemble households that share with their neighbors as opposed to households that use their connections individually.

@@

Let the city of Manila be partitioned into neighborhoods, where each neighborhood is defined by a set of households that are physically close enough to share water connections with each other.  In the context of Manila, these neighborhoods are often characterized by dense clusters of houses or duplexes.  This partition embeds the assumption that neighborhoods are mutually exclusive so that a household located in one neighborhood cannot share water with households in a nearby neighborhood.  Restricting the choice set in this way may tend to overestimate the costs that households face in finding a sharing partner.\footnote{This limitation is discussed in more detail in Section~\ref{section:fixedcostestimation}}

In the first period, each household makes a one-time decision about their water source between purchasing a connection, using a neighbor's connection, or buying from a local water vendor.  These decisions result in a set of water connections for each neighborhood ranging from an empty set if all households use from a vendor to equaling the size of the neighborhood if each household uses an individual connection.  If households choose to connect to the network, each connection, $C$, serves a set of households given by $S$, which depends on which households have decided to share the water connection.  Also, let the index $C$ indicate connection-level attributes, summing over all households using connection $C$.  This model does not allow for households to choose different water sources over time, which is consistent with large fixed costs involved in initially purchasing a connection or establishing a sharing relationship.  At the same time, some households may exogenously lose access to their water source, forcing these households to substitute toward other sources as detailed in Section~\ref{section:hasslemodel}.

Conditional on their water source choice, households then choose how much water to use each month.  Since water sources provide utility through monthly consumption, this section describes the decision process in reverse order, first characterizing monthly water consumption choices (taking the initial source choice as given) and second, describing the water source choice (according to the expected utilities from using each source).

@@

In choosing how much water to use each month, households using piped water face the difficult task of coordinating water consumption among all household members.  For example, consider a household head trying to manage water consumption across all of their household members each month.  This task would involve monitoring the meter located outside of their dwelling and observing water use among all members.\footnote{Given that many developing countries feature steeply increasing prices with monthly usage, these coordination challenges may magnify any distortions caused by non-linear pricing.}  Mobile survey data indicate that \input{tables/no_info}\unskip\% of surveyed household members do not recall how much they paid for water in the last month while \input{tables/no_use}\unskip\% do not recall the amount that their household consumed in the previous month.

As a simple way to capture this coordination task, let each household $i$ set a goal for water use each month, $w_i$, knowing that consumption shocks, $\epsilon_i$, may increase or decrease the realized amount of usage.  Consumption shocks, $\epsilon_i$, are assumed to be unobserved to the household and independent across time, bringing the total realized water consumption each month to, $W_i=w_i+\epsilon_i$.  Household utility over water, $W_i$ and a numeraire good, $x_i$ is assumed to take the following shape:
\begin{align}\label{equation:umaxfirst}
\begin{split}
U(x_i,W_i) &=  x_i + (1-H_i) W_i - \frac{1}{2\alpha_i } ( W_i - \Gamma_i + \alpha_i )^2  \\
W_i &= w_i + \epsilon_i \\
\epsilon_i &\sim \mathcal{N}(0,\Sigma_{\epsilon})
\end{split}
\end{align}
Following from quasi-linear preferences, this quadratic form for utility eliminates income effects and includes a natural satiation point for water consumption, capturing the tendency of households to target specific consumption levels despite low or zero marginal prices as in Figure~\ref{figure:consumptionhistogram}.  Household preferences for water are characterized by $\Gamma_i$, which can be interpreted as a fixed preference for water each month and $\alpha_i$, which captures the extent to which households are price sensitive.  These terms have clearer interpretations in the demand function below.  The exogenous shock to consumption, $\epsilon_i$, is assumed to be drawn from a multivariate normal distribution, $\Sigma_{\epsilon}$ with zero mean, variance terms, $\sigma_{\epsilon}^2$, and allowing for correlation with neighboring households, ${\rho}^{\epsilon}_{i,j}$.  Although households cannot anticipate the exact realization of $\epsilon_i$, they are assumed to understand the distribution of possible shocks, allowing them to maximize expected utility.

Households that are using water from a neighbor's tap may face an additional utility cost per gallon of water $H_i$.  This hassle cost captures any inconvenience associated with using water from a neighbor's tap.  The mobile survey reveals that among households accessing water from a neighbor's connection, \input{tables/access_fetch}\unskip\% fetch water with containers, \input{tables/access_single_tap}\unskip\% access water from a single hose/pipe provided by the connection owner, and \input{tables/access_pipes}\unskip\% are able to fully connect their household plumbing near the water meter.  These different modes of access suggest that sharing households may face higher time costs fetching water or lower water pressure from piping water additional distances.  Without data on the sharing mode and distance between households for each connection, the model assumes all sharing households face a homogeneous hassle cost, $H_i$, from using a neighbor's connection.\footnote{Section~\ref{section:hasslecostestimation} discusses the extent to which this assumption may bias estimates.  The partitioning of the city into sharing neighborhoods also imposes the implicit assumption that hassle costs rise substantially between neighborhoods so that no household has an incentive to share with households in another neighborhood.}

@@

The household problem is additionally complicated by the presence of non-linear price schedules for water.  Figure~\ref{figure:tariff} provides the increasing tariff in Manila.  The non-linear price schedule produces kinks in household budget constraints leading to non-linear demand functions.  Let $P(W_C) = \{ p^{1},...,p^{K},\overline{W}^{0},...,\overline{W}^{K}\}$ describe tariff schedule as a function of total consumption for the water connection, $W_C = \sum_{i=1}^{S} W_i$.  The tariff includes $K$ segments where segment $k$ corresponds to price $p^{k}$ for water consumption between the consumption thresholds (or kink points), $\overline{W}^{k-1}$ and $\overline{W}^{k}$:
\begin{align*}
P(W_C)=&  
\begin{cases}
      p^{1} &\text{ if } W_C \in [\overline{W}^{0}(=0),\overline{W}^{1}], \\
      p^{2} &\text{ if } W_C \in [\overline{W}^{1},\overline{W}^{2}], \\
      ...   &\text{ ...}, \\
      p^{K} &\text{ if } W_C \in[\overline{W}^{K-1},\overline{W}^{K}(=\infty)]
    \end{cases}
\end{align*}
Given prices and household utility, the final ingredient for characterizing consumption choices is the budget constraint.  The budget constraint includes monthly household income, $Y_i$, and monthly fixed costs of using the water connection, $F_i$, which are assumed to capture three components: (1) monthly service fees, (2) initial connection fees amortized in monthly terms for the duration of the connection, and (3) unobserved, fixed hassle costs associated with setting up a connection, which are also amortized into monthly terms.  Data does not track household water source choices over time in a comprehensive way, which limits the extent to which the model can distinguish between one-time connection costs and monthly service/maintenance costs.  This limitation means that the model cannot address the optimal timing of fixed payments or capture any potential heterogeneity in monthly fixed costs.

Water sharing relationships impact monthly consumption choices through the budget constraint.\footnote{While the formation of these relationships is discussed in more detail in Section~\ref{section:watersourcechoice}, this section takes these sharing relationships as exogenously given.}  To develop notation for sharing, normalize $i=1$ to be the owner of the connection and define the total number of sharing households for a particular connection as $S$.  Therefore, all households, $i\neq1$, using from a neighbor's tap have a positive hassle cost, $H$, while the connection owner, $i=1$, has zero hassle cost.

>>>>

%  Shared connections record high levels of monthly consumption, which can expose them to high prices when marginal prices increase with usage . 
This paper analyzes the welfare impacts of pricing policies in the presence of informal water sharing.  Welfare impacts depend not only on heterogeneity in demand for water across the population, but also on costs associated with forming and maintaining water sharing relationships.  To measure these features, I merge administrative data covering 1.5 million water connections in Manila from 2010 to 2015 with detailed survey data on the number of households and people using each water connection.  With comprehensive geographic coverage of half of Manila and a variety of price variation, these data allow for a rich characterization of water demand.  Additional variation in water source choices identify the relative fixed and marginal costs associated with sharing water with a neighbor.  I find that despite relatively high marginal costs of sharing, shared connections allow the city to save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households at low marginal prices.  At the same time, informal sharing networks serve as resale markets for water, undermining the extent to which cities can redistribute surplus through increasing marginal prices.

@@

Combining the model and estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,wsup,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensuring that the water provider exactly covers costs.  This policy reduces access by incentivizing households to substitute toward water vendors, which mainly include deepwells and water refilling stations.

@@

In focusing on water sharing behavior, this approach abstracts away from several features that may also affect optimal water pricing.  First, I do not directly model environmental, health, or other externalities associated with piped water use.  Water shortages may lead policymakers to promote conservation through higher marginal prices while positive health externalities may motivate further subsidizing piped water (\cite{timmins2002measuring}, \cite{galiani2005water}).\footnote{\cite{bennett2012does} shows that piped water can also lead to negative externalities through worse sanitation, possibly motivating higher prices for piped water in some cases.}  Second, I assume that government regulator is able to fully observe production costs for the water provider, eliminating scope for prices to correct for possible principal-agent frictions as characterized by \cite{laffont1993theory}.  Finally, the counterfactual exercises borrow the public-private partnership regulatory framework in Manila by assuming that revenue must fully cover production costs.  Additional data on the costs of raising public funds and risks of regulatory capture would allow for a more flexible analysis taking into account the possibility of public funding for water (\cite{laffont2005regulation}).

@@

%Monthly quantity choices across different households and at different prices reveal information about household demand.  Observing multiple quantities choices for the same household allows for a rich characterization of demand heterogeneity at the household-level.  
Evaluating different pricing policies requires measuring demand at the household level as well as costs associated with providing piped water.  Data was gathered through a partnership with one of the two regulated, private providers in Manila who provided both billing records as well as information about the regulatory structure, pricing, and production costs.  Complete billing records for each water connection including the monthly meter reading, the total bill charged, and the precise location for each meter serve as the primary measure of quantity choices.  These data span 2010 to 2015, during which connections grew from 900,000 to 1.5 million water connections, generating nearly 100 million transaction records.  Connections are divided into residential (90\%), semi-business (4\%), commercial (5\%), and industrial (1\%).  This paper focuses on residential and semi-business connections.\footnote{Semi-business connections are largely composed of households that own small roadside stands.  These connections provide a useful source of price variation which is described in Appendix~\ref{appendix:ratechangeelasticity}.}  The monthly meter reading is used as the measure of consumption for each connection.  Meter readings are rounded to the nearest cubic meter limited to monthly readings below 100 $\text{m}^{3}$ in order to exclude pipe leaks as well as meter reading errors (readings above 100 $\text{m}^{3}$ compose less than 0.7\% of the sample).

Since the object of interest for welfare is household-level demand, the billing records are merged to survey data recording information on households using each connection as well as demographics for the owners of each connection.  To evaluate the performance of the water provider, the regulator contracted with a local university to conduct the Public Assessment of Water Services survey in three waves from 2008 to 2012, surveying around 50,000 water connections.\footnote{Appendix~\ref{appendix:pawssampling} describes the sampling design and tests the extent to which it may be correlated with demographic characteristics.  This exercise finds relatively weak correlations in economic terms with key characteristics.}  The connection survey records the total number of households and people using each connection, providing the primary measure of sharing behavior.  The connection survey also includes basic demographics for the owner of the connection.  The demographics used in the analysis include household size, dwelling type (whether apartment, duplex, or single house), number of employed household members, and whether the head of household is employed in a low-skill profession.
%consumption and water source choices at the household-level.  
%To examine both consumption decisions and water source choices at the household-level, I construct a dataset that merges (1) monthly water usage for around 50,000 connections, (2) demographics and water sharing measures for households using these connections, and (3) a cross-section of all households and their water source choices.  Put together, these data form a representative sample of households in half of Metro Manila.

A full analysis of water usage in Manila also requires measuring households that are unconnected to the bulk water system and instead use water from local vendors.  The 2010 Census of Population and Housing records demographics for households that are using water from alternative sources.  Households are defined as connected to the piped network if they report using a ``faucet community water system'' for their cooking, laundry, or bathing needs.\footnote{The census was designed largely for rural populations so it does not directly ask whether households are connected to a piped water network.}  Alternatively, households are considered to be using from a water vendor if they report any other category, which includes deep wells, peddlers, or other sources.\footnote{Since demographic questions in the census and connection surveys are worded differently, Table~\ref{table:pawstocensus} of Appendix~\ref{appendix:pawstocensus} compares measures across the two data sources for households that own connections finding broadly similar patterns.}  Census data is merged geographically with the connection survey for 599 wards in Manila.

%To construct a profile of water source choices for the full population, I randomly sample from the pool of households using from vendors in 599 wards of the city so that the data form a representative sample when combined with households covered by the connection survey in each small area.
%\footnote{Regions are chosen to overlap with ward administrative/political boundaries while ensuring that I am able to obtain a large enough sample of vendor households for each area to perform the counterfactual analysis in Section~\ref{section:fixedcostestimation}.}

For additional information on water sharing relationships between households, I fielded a small, non-representative survey of 600 households via mobile phone application.  This survey asks about issues associated with using different water sources, basic demographics, as well as how the bill and connection fee are paid for sharing households.  Appendix~\ref{appendix:pollfish} compares this survey to the connection survey finding roughly similar rates of sharing and demographic characteristics.

%includes additional details on this survey and finds roughly similar measures between the connection survey and the mobile survey, suggesting that the mobile survey may provide a reasonable descriptive picture of sharing in Manila.

The water provider also maintains call records of all service complaints, which can be linked to individual connections.  This data is used to measure instances where households face unanticipated shocks to their water source, which are used to identify the marginal costs of using water from a neighbor.  Finally, the water provider tracks data on the marginal cost of pumping an additional cubic meter of water as well as the costs associated with installing new connections.  These data provide inputs into the provider's budget constraint, which are necessary for computing counterfactual pricing regimes.

Supplemental survey data from a subset of Manila include income measures that are used to approximate household income using the full set of demographic measures and their interactions.

@@

The welfare consequences of informal sharing networks depend on both the types of households that choose to share with neighbors as well as the costs of these networks.  To provide some descriptive evidence, Table~\ref{descriptives} merges the connection survey of water connections with billing records associated with each connection.  The columns group connections by the number of households they serve: one, two, or at least three.  Demographics for connection owners indicate that smaller, slightly younger, and poorer households are more likely to share water with neighbors.  Sharing households are over twice as likely to reside in duplexes possibly because transporting water between neighbors requires less effort for duplex residents.

Larger numbers of households sharing a water connection are correlated with more total water usage, but less water per person.\footnote{Total People Served by a connection includes both the owning household and any other households using a connection.}  This decline may have many possible drivers: (1) sharing households face higher prices by locating at a higher point on the tariff schedule (Figure~\ref{figure:tariff}), (2) sharing households may face hassle costs in fetching water from neighboring taps, (3) these price effects may incentivize low-demand households to select into sharing relationships, consistent with the demographic patterns, and (4) low-demand households may have idiosyncratic preferences for sharing relationships.  On average, monthly water bills (inclusive of service fees) take up around \input{tables/incshr}\unskip\% of household income, which does not include other unobserved costs in accessing water.  These findings suggest that policies which affect sharing choices can have significant impacts on water consumption as well as monthly expenditures.

%Demographics for connection owners indicate that smaller and younger households are more likely to share water with neighbors.  Because household size is a strong proxy for water demand, this finding suggests that sharing a connection is negatively correlated with household water demand.  Households with lower imputed income more commonly share water connections.  Sharing households are over twice as likely to reside in duplexes possibly because transporting water between neighbors requires less effort for duplex residents.

@@

%Table~\ref{descriptives} summarizes water use and sharing behavior for these connections, merging information from the connection survey with the billing records.

%Also, the mobile survey indicates that only around 30\% of sharing households would substitute to toward a neighbor's connection if they no longer had access to their primary water source.  For this reason, I treat connections reported to serve more than three households as only serving three households.  
Table~\ref{descriptives} also computes the total number of households using from each source.  This calculation requires the additional assumption that households sharing with one or two other households are uniquely linked to one connection.  This way, the population of households connected to the network can be computed by summing the number of households across all connections.  This assumption is consistent with the presence of fixed costs in setting up a connection or a sharing relationship so that households stay committed to one water source.  To the extent that households are drawing water from multiple connections, this assumption may lead to an overestimation of the population engaged in sharing relationships.  With this assumption, \input{tables/shr_individual}\unskip\% of connected households receive their water from an individual connection, \input{tables/shr_1hh}\unskip\% are involved in a sharing relationship with a single neighbor, and \input{tables/shr_2hh}\unskip\% are in a sharing relationship with 2 or more other households.  These proportions do not take into consideration households that are using from water vendors, which are discussed in more detail below.  Each water connection contributes over four years of billing records on average.

Since the connection survey only includes households connected to piped water, census of population and housing data is required to learn about households using from water vendors.  In this data, \input{tables/vendor_total}\unskip\% of the population report using vendors, which is composed of \input{tables/vendor_well}\unskip\% of households using from a ``deep well'' and \input{tables/vendor_peddler}\unskip\% using from a ``peddler'' while the remaining fall into an ``other'' category.  Anecdotal evidence suggests that groundwater from wells often has very low quality.  Appendix Table~\ref{table:pawstocensus} compares means of demographic variables between connection owners from the connection survey and all households from the census.  Demographic variables are similar between connection owners in the connection survey and the census data.  Households using from vendors have lower imputed incomes, greater probability of low-skill employment, and slightly fewer members than own-use households.  Across most dimensions, households using from vendors more closely resemble households that share with their neighbors as opposed to households that use their connections individually.

@@

Let the city of Manila be partitioned into neighborhoods, where each neighborhood is defined by a set of households that are physically close enough to share water connections with each other.  In the context of Manila, these neighborhoods are often characterized by dense clusters of houses or duplexes.  This partition embeds the assumption that neighborhoods are mutually exclusive so that a household located in one neighborhood cannot share water with households in a nearby neighborhood.  Restricting the choice set in this way is necessary for computation but may in some cases overestimate the costs that households face in finding a sharing partner.\footnote{This limitation is discussed in more detail in Section~\ref{section:fixedcostestimation}}

In the first step, each household makes a one-time decision about their water source between 1) purchasing a connection, 2) using a neighbor's connection, or 3) buying from a local water vendor.  These decisions result in a set of water connections for each neighborhood ranging from an empty set (if all households use from a vendor) to the whole neighborhood (if each household uses an individual connection).  Let each connection, $C$, serve $S$ households that have decided to share the connection.  For ease of notation, let the subscript $C$ indicate connection-level attributes, summing over all households using from connection $C$.

This model does not allow for households to choose different water sources over time, which is consistent with large fixed costs involved in initially purchasing a connection or establishing a sharing relationship.  Section~\ref{section:hasslemodel} extends the model to allow for cases where households exogenously lose access to their current water source and must substitute toward another nearby source.

In the second step, each household then chooses how much water to use each month from their source.  Since water sources provide utility through monthly consumption, the following section describes the decision process in reverse order, first characterizing monthly water consumption choices (taking the initial source choice as given) and second, describing the initial water source choices.

@@

In choosing how much water to use each month, households using piped water face the difficult task of coordinating water consumption among all household members.  For example, consider the head of a household trying to manage water consumption across all of their household members each month.  This task would involve monitoring the meter located outside of their dwelling and coordinating water use among all members.\footnote{Given that many developing countries feature steeply increasing prices with monthly usage, these coordination challenges may magnify any distortions caused by non-linear pricing.}  Consistent with these challenges, mobile survey data indicate that \input{tables/no_info}\unskip\% of surveyed household members do not recall how much they paid for water in the last month while \input{tables/no_use}\unskip\% do not recall the amount that their household consumed in the previous month.

As a simple way to capture this coordination task, let each household $i$ set a goal for water use each month, $w_i$, knowing that unobserved consumption shocks, $\epsilon_i$, may increase or decrease their realized amount of usage.  Consumption shocks, $\epsilon_i$, are assumed to be unobserved to the household and independent across time, bringing the total realized water consumption each month to, $W_i=w_i+\epsilon_i$.  Household utility over total water, $W_i$ and a numeraire good, $x_i$ is assumed to take the following shape:
\begin{align}\label{equation:umaxfirst}
\begin{split}
U(x_i,W_i) &=  x_i + (1-H_i) W_i - \frac{1}{2\alpha_i } ( W_i - \Gamma_i + \alpha_i )^2  \\
W_i &= w_i + \epsilon_i \\
\epsilon_i &\sim \mathcal{N}(0,\Sigma_{\epsilon})
\end{split}
\end{align}
Following from quasi-linear preferences, this quadratic form for utility eliminates income effects and includes a natural satiation point for water consumption, capturing the tendency of households to target specific consumption levels despite low or zero marginal prices as in Figure~\ref{figure:consumptionhistogram}.\footnote{This utility form builds on \cite{einav2013selection} who use a similar form to characterize extensive and intensive choices of health insurance.}   Household preferences for water are characterized by $\Gamma_i$, which can be interpreted as a fixed preference for water each month and $\alpha_i$, which captures the extent to which households are price sensitive.  The exogenous shock to consumption, $\epsilon_i$, is assumed to be drawn from a multivariate normal distribution, $\Sigma_{\epsilon}$ with zero mean and variance terms, $\sigma_{\epsilon}^2$, allowing for correlation with neighboring households $i$ and $j$ through term ${\rho}^{\epsilon}_{i,j}$.  Although households cannot anticipate the exact realization of $\epsilon_i$, they are assumed to understand the distribution of possible consumption shocks, allowing them to maximize expected utility.

Households that are using water from a neighbor's tap may face an additional utility cost per gallon of water $H_i$.  This hassle cost captures any inconvenience associated with using an additional unit of water from a neighbor's tap.\footnote{Accordingly, the connection owner is assumed to face zero hassle cost.}  The mobile survey reveals that among households accessing water from a neighbor's connection, \input{tables/access_fetch}\unskip\% fetch water with containers, \input{tables/access_single_tap}\unskip\% access water from a single hose/pipe provided by the connection owner, and \input{tables/access_pipes}\unskip\% are able to fully connect their household plumbing near the water meter.  These different modes of access suggest that sharing households may face higher time costs fetching water or lower water pressure from piping water additional distances.

%To develop notation for sharing, normalize $i=1$ to be the owner of the connection and define the total number of sharing households for a particular connection as $S$.  Therefore, all households, $i\neq1$, using from a neighbor's tap have a positive hassle cost, $H$, while the connection owner, $i=1$, has zero hassle cost.

%FIX HERE!

%Without data on the sharing mode and distance between households for each connection, the model assumes all sharing households face a homogeneous hassle cost, $H_i$, from using a neighbor's connection.\footnote{Section~\ref{section:hasslecostestimation} discusses the extent to which this assumption may bias estimates.  The partitioning of the city into sharing neighborhoods also imposes the implicit assumption that hassle costs rise substantially between neighborhoods so that no household has an incentive to share with households in another neighborhood.}

@@

The household problem is additionally complicated by the presence of non-linear price schedules for water.  Figure~\ref{figure:tariff} provides the increasing tariff in Manila.  The non-linear price schedule produces kinks in household budget constraints leading to non-linear demand functions.  Let $P(W_C) = \{ p^{1},...,p^{K},\overline{W}^{0},...,\overline{W}^{K}\}$ describe tariff schedule as a function of total consumption for the water connection, $W_C = \sum_{i=1}^{S} W_i$.  The tariff includes $K$ segments where segment $k$ corresponds to price $p^{k}$ for water consumption between the consumption thresholds (or kink points), $\overline{W}^{k-1}$ and $\overline{W}^{k}$:\footnote{In this case in Manila, $\overline{W}^{0}$ is zero while $\overline{W}^{K}$ is infinity.}
\begin{align*}
P(W_C)=&  
\begin{cases}
      p^{1} &\text{ if } W_C \in [\overline{W}^{0},\overline{W}^{1}], \\
      p^{2} &\text{ if } W_C \in [\overline{W}^{1},\overline{W}^{2}], \\
      ...   &\text{ ...}, \\
      p^{K} &\text{ if } W_C \in[\overline{W}^{K-1},\overline{W}^{K}]
    \end{cases}
\end{align*}
Given prices and household utility, the final ingredient for characterizing consumption choices is the budget constraint.  The budget constraint includes monthly household income, $Y_i$, and monthly fixed costs of using the water connection, $F_i$, which are assumed to capture three components: (1) monthly service fees, (2) initial connection fees amortized in monthly terms for the duration of the connection, and (3) unobserved, fixed hassle costs associated with setting up a connection, which are also amortized into monthly terms.  Data does not track household water source choices over time in a comprehensive way, which limits the extent to which the model can distinguish between one-time connection costs and monthly service/maintenance costs.  This limitation means that the model cannot address the optimal timing of fixed payments or capture potential heterogeneity in monthly fixed costs.
